### 2020-12-01 paper/2020-10-revision-background.tex
First complete draft of conceptual framework

This section provides a conceptual framework for understanding how birth spacing changes
and guiding the subsequent analysis.
In non-Western countries, lower fertility is associated with longer birth spacing, but 
outside a ``natural fertility'' regime it is not clear that lower fertility necessarily 
leads to longer spacing \citet{Newman1988,Rutstein2011,Casterline2016}.
Thus, I first introduce three potential explanations that link fertility and birth 
spacing decisions: economic conditions, desired child outcomes, and son preference 
\citep{Casterline2016,Portner2018}.
Next, I argue that female education and area of residence are the most important factors 
to consider in the empirical analysis and discuss how they tie into the three theories.
Finally, I describe how female education, together with female labor force participation, 
have changed over time in India and how these changes affect the predictions for how birth 
spacing has changed.

@@

% Factors/predictions
An important issue is how to operationalize the potential explanations of how 
birth spacing has changed over time.
Although the explanations are based on changes in wages, mortality risk, degree of son 
preference, and access to and cost of sex selection, none of these are available for the 
period when the decisions of interest were made.

Two characteristics that are available in data and that have been shown to be associated 
with fertility, child outcomes, and the use of sex selection are female education and 
area of residence.
[citations?]
The problem is that both have multiple pathways through which they can affect fertility,
birth spacing, and sex selection and that these do not necessarily line up neatly with the
theories above.

@@

Wages are generally increasing in education suggesting that with higher female education
we should see shorter birth spacing and lower fertility, but also higher use of sex 
selection.
However, female education undoubtedly represents more than the value of their time.
Two examples are child health and the production of offspring human capital.

First, children of more educated women tend to have better health outcomes and
lower mortality than less educated women, which may come from higher income, but could
also be a result of better health knowledge 
\citep{Rosenzweig1982a,Kovsted2002,Whitworth2002,Maitra2008}.
Lower expected mortality means that fewer births are required to reach the desired
number of surviving children.
Similarly, although short birth spacing generally leads to worse health outcomes, 
the more educated the mother, the more muted this relationship is, which, at least partly,
removes the health outcome incentive for longer birth spacing
\citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

Second, more educated women are more efficient at producing child human capital 
\citep{Behrman1999}.
Hence, families that value sons' education more will both tend to demand better-educated
women as mothers and to have fewer children because both allow for higher quality boys.
With the increasing return to male education that India has experienced, this imply
that there will be an increased demand for better-educated women combined with lower 
fertility, and an increased use of sex selection when available.
In this case, the associations between higher female education, lower fertility, and more
sex selection are not a causal effect driven by higher wages but instead the result of the 
increased return to male education.

% Descriptive

[move to the prediction part?]
In India, real wages for both men and women have almost doubled between 1987 and 
2011, although the mean male wage is still close to 70\% higher than the female wage 
\citep{Klasen2015,Bhargava2018}.

>>>>

This section provides a conceptual framework for understanding how birth spacing changes
and guiding the subsequent analysis.
In non-Western countries, lower fertility is associated with longer birth spacing, but 
outside a ``natural fertility'' regime it is not clear that lower fertility necessarily 
leads to longer spacing \citet{Newman1988,Rutstein2011,Casterline2016}.
Thus, I first introduce three potential explanations that link fertility and birth 
spacing decisions: economic conditions, desired child outcomes, and son preference 
\citep{Casterline2016,Portner2018}.
Next, I argue that female education and area of residence are the most important factors 
to consider in the empirical analysis and discuss how they tie into the three theories.

@@

% Factors/predictions
Female education and area of residence are the two principal explanatory variables I use
in the empirical analyses.
The main reason is that female education and area of residence are the closest proxies in 
the data for wages and access to and cost of sex selection, none of which are available.
In addition, both higher female education and urbanization are associated with lower 
fertility and increased use of sex selection in past research
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.

Finally, India has seen tremendous increases in both education and urbanization.
The proportion of the population living in urban areas almost doubled from 18\%
in 1960 to 35\% in 2019 \citep{United-Nations2019}.
Figure \ref{fig:education_over_time} shows the distribution of schooling by birth 
cohort for urban and rural women, 20 years or older, whether married or not, in the 
four rounds of the NFHS.
Female education has increased substantially in both rural and urban areas,
but is still considerably higher in urban than in rural areas.
In rural areas, women with eight or more years of education have gone from almost zero for 
the 1930 cohort to more than 60\% for the 1990s cohorts, with about half in the 8--11 
group and the other half in the 12 plus group.
In urban areas, most of the increase in urban female education came from the 12 plus 
group, which now accounts for more than half of all urban women for the most recent cohorts.

\begin{figure}[htpb]
\centering
\subfloat[Rural]{
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_rural}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_urban} 
    \end{minipage}
}
\caption{Distribution of education by cohort for women 20 years or older at survey}
\label{fig:education_over_time}
\end{figure}

@@

Wages generally increase with education suggesting that higher female education
should shorten birth spacing and lower fertility, but also increase the use of sex 
selection.
However, female education undoubtedly represents more than the value of women's time.
Two examples are the productions of child health and offspring human capital.
[something on norms?]

First, children of more educated women tend to have better health outcomes and lower 
mortality than less educated women, either from higher income or better health knowledge 
\citep{Rosenzweig1982a,Kovsted2002,Whitworth2002,Maitra2008}.
Lower expected mortality means that fewer births are required to reach a desired
number of surviving children.
Similarly, although short birth spacing generally leads to worse health outcomes, 
the more educated the mother, the more muted this relationship is, which, at least partly,
removes the health outcome incentive for longer birth spacing
\citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

Second, better-educated women are more efficient at producing child human capital 
\citep{Behrman1999}.
[need something on son preference? Maybe simply use increasing returns to male education]
Hence, families that value sons' education more will tend to both demand better-educated
women as mothers and have fewer children because both allow for higher quality boys.
With the increasing return to male education that India has experienced, this imply
an increased demand for better-educated women combined with lower fertility, and an 
increased use of sex selection when available.
In this case, the associations between higher female education, lower fertility, and more
sex selection are not a causal effect driven by higher wages but instead the result of the 
increased return to male education.

Even as the level of female education has increased, the female labor force 
participation in both urban and rural areas has stagnated or decreased
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
India's female labor force participation is now one of the lowest in the world  
and does not yet show any signs of increasing \citep{Klasen2015,Chatterjee2018}.
In line with previous research, the NFHS data show a U-shaped relationship between 
education and working for married women, with the highest percentage working being women 
with either no education or 12 or more years and the lowest being women with 8--11 years 
of education.%
\footnote{
See appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}

The low and declining female labor force participation, especially for younger women, 
suggests that families face little incentive to space children more closely 
together for economic reasons.
One explanation is that household income has increased so substantially that the income 
effect dominates any substitution effect.
Two findings speak to this effect.
First, although real wages for both men and women have almost doubled between 1987 and 
2011, the mean male wage is still close to 70\% higher than the female wage 
\citep{Klasen2015,Bhargava2018}.
Second, women’s labor supply appears to be more negatively elastic to husbands' 
wages than positively elastic to their own wages \citep{Bhargava2018}.

% Sanskritization

The increases in female educational attainment imply that access to education has 
expanded beyond the higher castes. 
One possible effect of the associated change in the composition of better-educated women 
is that this group's behavior would change.
However, ``Sanskritization'' implies that as lower-castes females gain access to 
education and their husbands' income increases, they adopt higher-caste 
norms such as stronger son preference and a retraction from the formal labor 
market \citep{Srinivas1956,Chen1995,Abraham2013,Chatterjee2018}.
The low and declining female labor force participation suggests that this process still operates.

% Summary / hypotheses

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
Furthermore, I expect birth spacing to increase the most among the better educated 
because their household income increases the most---even with declining female labor 
participation---and because of their use of sex selection.
Even with the substantial increase in the number of better-educated women, 
"Sanskritization" implies that the changing composition will not substantially change 
the use of sex selection.
